Scalable Clustering of Genotype Information using MapReduce

### 2014-12-18 doc/publications/journal2014/paper.tex
Additions

\section{Motivation:}
Processing genomic information from whole genome sequence studies pose computational challenges due to the unprecedented data volume generated, which render transitional approaches insufficient. However, by utilising advancements in modern hardware accelerators and data processing we can provide the means for scalable solutions. We therefore aim to provide the interface between standard genomic data formats and advanced and scalable analysis libraries like Mahout. 
\section{Results:}
We achieve an 2-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R, by comparable accuracy. However, the real benefit lies in scaling beyond R's capability to a population-size analysis. We successfully clustered more than 5000 individuals each having more than 15 Million variants.

@@

Grouping individuals based on the genomic profile is a commonly performed tasks to identify population structure~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.  Traditionally both the number of individuals and included genotypes, typically from SNP arrays, were relatively small and libraries in Bioconductor sufficient. However, recent technological advances in whole genome sequencing have made population-scale sequencing feasible. It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or the cancer genome atlas (TCGA)~\cite{TCGA2013}. At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude. This drastic increase in both sample numbers and features per sample requires a massively parallele approach to data processing.

@@

To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure as well as ancestry using the k-means clustering algorithm available in Mahout. In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a reduced version of the data and in the last two sections we investigate how Mahout scales to either more samples or more features per samples.

@@

% COMPARISON TO R
\subsection*{Clustering in Mahout uses less memory than in R}
In this section, we compare the runtime between the k-means implementation in R and Mahout for the small \NinteenPhaseone\ dataset. 
To do this we measured the runtime separately for pre-processing stage as well as the k-means clustering stage itself (see Methods). 
Pre-processing the data to Mahout's sequence files takes 21 minutes, uses 1.8GB physical memory (pmem) and 7.6GB virtual memory (vmem).
Conversely,  pre-processing the data to R's in memory matrix takes 56 minutes, uses 17.7GB pmem and 18GB vmem. 
The size of the Mahout sequence file, which is saved to disk, is 835MB. 
K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes 10 minutes. See Table~\ref{datasetsTime}. \\
Due to the small number of classifying variants, cluster performance is very poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.769 for 20 clusters in Mahout and 0.614 with 10 clusters in R (see Table~\ref{datasetsAcc}).

Although the Mahout pipeline is about 25 minutes faster than R's, the actual clustering step takes two orders of magnitude longer. 
This is likely due to the additional overhead involved in setting up the Hadoop nodes, spawning the map and reduce instances, and distributing the data between the instances. 
R on the other hand, already had the data in memory, and only had to process that data. 
In the case of this comparison, the data set is likely too small to truly reap the benefit of Hadoop's MapReduce algorithm. 
The main benefit of Mahout in this scenario is the memory consumption with nearly 10-fold reduction for Mahout compared to R.

% SCALING UP FEATURES IN MAHOUT
\subsection*{Mahout scales to 15 Million genomic variants}
In this section we demonstrate Mahout's scaling ability by increasing the numbers of features, here variants per sample. 
Including more regions for the similarity calculation will increase the cluster performance, however also increases the memory consumption. 
R therefore was not able to perform the clustering on even the medium sized dataset (\SevenPhaseone , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
On this dataset Mahout's pre-processing takes 325 minutes, with 13GB of physical and 37GB of virtual memory, and the clustering takes 80 minutes.
The clustering accuracy has increased by 35\% to 0.812 with 10 clusters compared to the performance on \NinteenPhaseone\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).

Utilizing the whole dataset hence would likely increased the performance further. However, on \FullPhasethree\ (38 Million variants) Hadoop returns an 'out-of-memory' error on the name-node before the jobs can even be distributed to the mappers. 
%TODO: is this a correct statement?
It is not possible to use Mahout's standard implementation for this dataset, as the supported Hadoop-setup only allows the communication between nodes with up to 8GB of memory and a scaling beyond that would have been required.

While R does not scale to either of the two orders of magnitude incased datasets, Mahout copes will with up to 15 Million variants. Going beyond this would require an implementation that also paralizes the individual variants as memory directly scales with the size of this feature vector.

% SCALING UP SAMPLES IN MAHOUT
\subsection*{Mahout scales effortless to an order of magnitude more samples}
In this section we investigate Mahout's ability to scale with more samples.
The resource consumption remains virtually identical between \SevenPhaseone\ and the order of magnitude larger dataset \OnePhasethree, which takes 33 minutes, with 6GB of physical and 39GB of virtual memory, and the clustering takes 23 minutes 
The clustering accuracy is 0.812 with 10 clusters.

@@

\section*{Conclusion}
In this paper we have applied Mahout's k-means clustering to the task of genotype based population and family association. 
We provide an interface from VCF format to Mahout and a toolset to interpret Mahout's clustering output (see Figure 1)
We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (10 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes. 
However, increasing either the number of samples or the number of features per sample quickly exceeds the ability of R.
Mahout on the other hand scales effortlessly to five times more samples (5,460) and copes well with 15 Million features per samples.
As memory consumption directly scales with the number of features per sample, we observe that Mahout's standard set-up of Hadoop does not scale to a whole exome dataset (38 Million features per sample).
However, this might be resolved once Mahout moves from MapReduce to a domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}.

@@

\begin{methods}
\section{Methods}
\subsection*{Datasets}
We used variant call format (VCF) files from the 1000 Genomes Project.
These VCF files contain information about the genetic variants between each of the 1092 individuals and a reference genome derived from GRCh37. 
Each of the VCF files are partitioned by chromosome.
Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
To be able to compare with R, we created a subset of the entire human variant dataset by using only variants from chromosome 19 (\NinteenPhaseone ) where we investigate the correct grouping of the population structure. 
Scaling up using Mahout, we investigate how the accuracy improves when using more variants from the genome with \SevenPhaseone\ and \FullPhasethree, which uses chromosome 1-5 in addition to 19 and the full genome, respectively. 
We also investigate whether we can accurately predict the more detailed family association using these datasets. 
In addition, we demonstrate Mahout's scalability by increasing the number of samples in the study to 5,460 by duplicating the original 1092 samples 5 times with zero and 10\% noise levels, i.e. randomly changing 10\% of the genotypes. 
Table~\ref{datasets} contains an overview of the datasets.

@@

\subsection*{Pre-processing}
\label{Sec:preprocessing}
Mahout requires Hadoop's `SequenceFiles' as input to its machine learning algorithms. SequenceFiles are binary flat files consisting of key/value pairs. In this case, each of the keys is a
label for an individual. The associated value for each key is a `sparse vector' representing the individual's genomic variants. We use sparse vectors, rather than
dense vectors, to significantly reduce processing time and required disk space. Each sparse vector consists of two separate vectors. An item in the first vector is
effectively a key for an item at the same position in the send vector. The key represents a genomic location, where its value represents the variant. Zero values (positions where there is no variant)
are omitted. Therefore, the fewer variants in an individual's genome, the smaller the sparse vector size. In comparison, `dense vectors' are single arrays that
include every value. Therefore, the size of a dense vector isn't reduced with a large number of non-variants.

The values of the sparse vectors, which, as previously mentioned, represent an individuals genomic variants, are a \texttt{double} java type.
We set the value of each double to the Hamming distance between each variant and the reference genome. Therefore, a value identical to the reference will be 0 (and omitted from the sparse vector),
a heterozygous variant will be 1 and a homozygous variant will be 2.

We wrote a custom `MapReduce' class to convert VCF files to the aforementioned sequence files. This should work on any number of files,
of any file size; limited only by disk space and time. The `map' step creates key value pairs where the primary key is an individual, the secondary key is a variant location, and the value is the variant.
We used a secondary key as this allowed us to implement custom functions to collect and sort the output of each mapper, based on this secondary key. This significantly reduced
the overall processing time. The `reduce' step collects these sorted key/value pairs by primary key (individual ID) and adds the secondary key (location) and value (variant) to a sparse array. The algorithm
saves these sparse arrays and keys to disk in the aforementioned SequenceFile format.

\subsection*{Mahout K-means clustering}
To cluster the samples from the sequence files, we use Mahout 0.9 in conjunction with Hadoop 2.5.0. 
The clustering steps are the same for each dataset, however, due to the increased feature size we designate more memory to the
YARN containers. For the first step, we specify \texttt{k},
the number of clusters. The function \texttt{RandomSeedGenerator.buildRandom} creates a new sequence file containing \texttt{k}
random centroids from the samples sequence file. The sequence file
of samples and the sequence file of centroids serve as the input to \texttt{KMeansDriver.run}, a static Mahout API function that launches
a k-means clustering job on the Hadoop service.

\subsection*{R K-means clustering}
We use `readVcf' from the R library, `VariantAnnotation', to iteratively read in the VCF file. For each iteration, we convert the data to a matrix and
change the values for each variant to the Hamming distance, as mentioned previously. We then transpose the matrix, so individuals are represented
by rows, and append the matrix to that from the previous iteration. Once the matrix contains all of the variants from the VCF file, we cluster the data
using the `kmeans' algorithm from the `stats' library.

@@

\subsection*{Performance}
We recorded the memory usage and run-time from the PBS logs. For the Mahout jobs, we queued separate jobs
for pre-processing and k-means clustering to get an accurate representation of the time and memory requirements for each stage.
For the R job, where we could not easily separate the pre-processing and clustering components, we printed the time between
stages using the function, \texttt{proc.time()}.

@@

\begin{table}[!t]
\processtable{Datasets used in this study.\label{Tab:01}}
{\begin{tabular}{lrrrrr}\toprule
& individuals  & variants  & population & families\\
& & &groups&& \\\midrule
        \NinteenPhaseone & 1,092 & 816,115  & 4 & 855\\
        \OnePhaseone & 1,092 & 3,007,196  & 4 & 855\\
        \SevenPhaseone & 1,092 & 18,984,880 & 4 & 855\\
	\OnePhasethree\ & 2,535 & 6,468,094 & 5 & 855\\
	\ThreePhasethree\ & 2,535 & 19,381,970 & 5 & 855\\
	\FullPhasethree\ & 2,535 & 79,000,000 & 5 & 855\\\botrule
\end{tabular}}{Datasets used in this study}
\end{table}

\begin{table}[!t]
\processtable{Adjusted Rand index (accuracy).\label{Tab:01}}
{\begin{tabular}{lcc}\toprule
Tool, Dataset & accuracy \\\midrule
      R, \NinteenPhaseone & 0.614  \\ 
        Mahout, \NinteenPhaseone & 0.730\\
        Mahout, \OnePhaseone & 0.891\\
        Mahout, \SevenPhaseone & 0.812 \\
        Mahout, \OnePhasethree & 0.877  \\\botrule
\end{tabular}}{Datasets used in this study}
\end{table}

\begin{table*}[!t]
\processtable{Resources consumed.\label{Tab:01}}
{\begin{tabular}{lccccc}\toprule
&&\multicolumn{2}{ c }{Memory (GB)}&\multicolumn{2}{ c }{Time} \\
& - & map & reduce  & pre-processing & clustering\\\midrule
         R, \NinteenPhaseone & - & 1.5 & 2 &3,341 & 7\\
        Mahout,\NinteenPhaseone & - & 1.5 & 2 & 2m 59s & 9m 12s\\
        Mahout,\OnePhaseone & - & 1.5 & 2 & 7m 24s & 23m 9s\\
        Mahout, \SevenPhaseone & - & 1.5 & 2 & 47m 43s & - \\ 
        Mahout, \SevenPhaseone & - & 2 & 4 & - & 4820 \\ 
        Mahout, \OnePhasethree & - & 1.5 & 2 & 27m 48s & 89m 42s \\
        Mahout, \ThreePhasethree & - & 1.5 & 2 & 82m 14s & - \\
        Mahout, \FullPhasethree & - & 1.5 & 2 & 336m 8s & - \\\botrule
\end{tabular}}{Datasets used in this study}
\end{table*}

>>>>

\section{Motivation:}
Processing genomic information from whole genome sequence studies pose computational challenges due to the unprecedented data volume generated, which render transitional approaches insufficient. However, by utilising advancements in modern hardware accelerators and data processing we can provide the means for scalable solutions. We therefore aim to provide the interface between standard genomic data formats and advanced and scalable analysis libraries like Mahout. 
\section{Results:}
We achieve an 2-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R, by comparable accuracy. However, the real benefit lies in scaling beyond R's capability to a population-size analysis. We successfully clustered more than 2,500 individuals each having more than 19 Million variants.

@@

Grouping individuals based on the genomic profile is a commonly performed tasks to identify population structure~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.  Traditionally both the number of individuals and included genotypes, typically from SNP arrays, were relatively small and libraries in Bioconductor sufficient. However, recent technological advances in whole genome sequencing have made population-scale sequencing feasible. It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or the cancer genome atlas (TCGA)~\cite{TCGA2013}. At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude. This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing.

@@

To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure using the k-means clustering algorithm available in Mahout. In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a reduced version of the data and in the last two sections we investigate how Mahout scales to either more samples or more features per samples.

@@

% COMPARISON TO R
\subsection*{Clustering in Mahout uses less memory than in R}
In this section, we compare the runtime between the k-means clustering implementation in R and Mahout for the small \NinteenPhaseone\ dataset (see Methods). 
We measured the runtime separately for pre-processing stage as well as the clustering stage. 
With Hadoop, pre-processing the data takes approximately 3 minutes, with an allocation of up to 2GB memory per container. The resulting data is stored as a Hadoop 'SequenceFile' of about 835MB in size.
Conversely,  pre-processing the data to R's in-memory matrix takes 56 minutes and uses 17.7GB memory. 
K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes just over 9 minutes. See Table~\ref{datasetsTime}. \\
Due to the relatively small number of variants, cluster performance is very poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.730 in Mahout and 0.614 in R (see Table~\ref{datasetsAcc}).

Although the overall Mahout pipeline is about 43 minutes faster than R's, Mahout's k-means clustering takes two orders of magnitude longer. 
This is likely due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
R, on the other hand, already has the entire matrix in-memory, on a single machine, this speeds up the clustering significantly as no disk-access or networking is required. However, the amount of memory available on a single machine can quickly becomes a limiting factor. As an example, the entire set of 1000 Genome Project Phase 3 variants would require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{datasetsAcc}).

Also, although Hadoop, in this case, has access to a total of 36GB of memory distributed over 3 machines, the container size is only 2GB. Therefore, we could
easily run the same job on a single 12GB node, at the expense of time.

This example demonstrates the potential for scalability in Hadoop, compared to single machine solutions, in respect to it's potential applications to genomics. We will proceed to demonstrate that Mahout can expand with relative ease to encompass larger data, although this depends on the task and how it can be parallelized.

% SCALING UP FEATURES IN MAHOUT
\subsection*{Hadoop scales to 19 Million genomic variants}
In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample).

R therefore was not able to perform the clustering on even the second smallest dataset (\OnePhaseone , 3 Million variants). This dataset, stored as a double-precision matrix would consume approximately 26GB memory, hence why the in-memory implementation of R fails. Therefore we will only
report the results for Hadoop.
On this dataset Hadoop's pre-processing takes 7 minutes, using the same container size as the smaller dataset. Clustering takes 23 minutes.
The clustering accuracy increased by 35\% to 0.891, compared to the performance on \NinteenPhaseone\ (see Table~\ref{datasetsAcc}).

To perform k-means clustering on the entire dataset in a reasonable amount of time would require a larger cluster with more memory per container.
This is because each Map task loads the entire feature vector for the clusters into memory.
Although we could increase the container size to the full 12GB per node, this would mean only 3 containers (1 per node) could run simultaneously, rather than the maximum of 18, and would greatly increase the processing time.
Pre-processing the data, however, does scale to the entire feature-set, without an increase in container memory. Having the data in the distributed
SequenceFile format opens the door to other distributed analysis techniques.

While R does not scale to either of the two orders of magnitude incased datasets, We have demonstrated Hadoop to cope well with up to 70 Million variants. Although k-means clustering with a feature-set of this size is not feasible on a compute cluster of this size, clustering a smaller subset of the genome is a trivial task.

% SCALING UP SAMPLES IN MAHOUT
\subsection*{Hadoop scales effortless to an order of magnitude more samples}
In this section we investigate Mahout's ability to scale with more samples, i.e. the 1092 individuals in the Phase 1 dataset and the 2535 individuals in the Phase 3 dataset.
We used the \SevenPhaseone\ and the \ThreePhasethree\ data-sets specifically because they both contain a similar number of variants (19 million).
The container size required for pre-processing the data is identical for both datasets, with the Phase 3 dataset requiring less than double the time of the smaller Phase 1 dataset.
In this case, due to the number of Feature Vectors increasing, rather than the feature size, the memory requirements for k-means clustering should remain similar, as the cluster centres loaded into the containers' memory are approximately the same size.
This proved to be the case, as 4GB allocated per container was the minimum for both datasets. The time to complete clustering was much higher though,
especially due to the fact that only 9 4GB containers can run simultaneously, rather than the 24 1.5GB containers. 
The clustering accuracy is 0.816 for the Phase 1 dataset and 0.xxx for the Phase 3 dataset.

@@

\section*{Conclusion}
In this paper we have applied Mahout's k-means clustering to the task of genotype based population. 
We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 1)
We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes. 
However, increasing either the number of samples or the number of features per sample quickly exceeds the ability of R.
Mahout on the other hand scales effortlessly to more than twice the number samples (2,535) and can process data with over 79 Million features per samples.
As the memory required for k-means clustering directly scales with the number of features per sample, we observe that Mahout's implementation
of k-means does not scale to the entire exome dataset, unless memory per container were increased beyond the limits of our cluster.
However, this might be resolved once Mahout moves from MapReduce to a domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}.

@@

\begin{methods}
\section{Methods}
\subsection*{Datasets}
We used variant call format (VCF) files from phase 1 and 3 of the 1000 Genomes Project.
The VCF files contain information about the genetic variants between each of the 1092 and 2,535 individuals respectively and a reference genome derived from GRCh37. 
Each of the VCF files contain the variants in one chromosome, and we only use the 22 autosomes.
Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
We used different subsets of the data for different comparisons. For example, to be able to compare Hadoop with R, we used a subset of the entire human variant dataset containing only chromosome 19 variants (\NinteenPhaseone ).
Scaling up using Hadoop, we investigate how the accuracy improves when using more variants from the genome with \SevenPhaseone\ and \FullPhasethree, which include chromosome 1-7 and the full genome, respectively.
Table~\ref{datasets} contains an overview of the datasets.

@@

\subsection*{Pre-processing}
\label{Sec:preprocessing}
Mahout requires Hadoop's `SequenceFiles' as input to its machine learning algorithms. SequenceFiles are binary flat files consisting of key/value pairs. In this case, each of the keys is a
label for an individual, where the associated value for each key is a `sparse vector' representing the individual's genomic variants. We use sparse vectors, rather than
dense vectors, to significantly reduce required memory and disk space. Each sparse vector consists of two separate vectors. An item in the first vector is
effectively a key for an item at the same position in the second vector. The key represents a genomic location, where its value represents the variant. Zero values (positions where there is no variant)
are omitted. Therefore, the fewer variants in an individual's genome, the smaller the sparse vector size. In comparison, `dense vectors' are single arrays that
include every position where a variant may fall, whether or not there actually is a variant. Therefore, the size of a dense vector isn't reduced with a large number of non-variants.

The values of the sparse vectors, which, as previously mentioned, represent an individuals genomic variants, are a \texttt{double} java type.
We set the value of each double to the sum of the square root of the Hamming distance between each variant allele and the reference genome. Therefore, a value identical to the reference, 0|0, will be 0 (and omitted from the sparse vector),
a heterozygous variant, 0|1 or 1|0, will be 1 (sqrt1) and a homozygous variant, 1|1, will be 2 (sqrt1 + sqrt1). The square root of the each allele prevents variants such as 1|1 colliding with the variant 2|0.

We wrote a custom `MapReduce' class in Java to convert VCF files to the aforementioned sequence files. This is scalable and should work on any number of files,
of any file size; limited only by disk space and time. The `map' step creates key value pairs where the primary key is an individual, the secondary key is a variant location, and the value is the variant.
We used a secondary key as this allowed us to implement custom 'collect' and 'sort' classes to process the output of each mapper, based on this secondary key.
This significantly reduced the overall processing time.
The `reduce' step processes the key/value pairs, which were collected by their primary key (individual ID), and sorted by their secondary key (location), and  adds the values (variants) to a sparse array.
The algorithm saves these sparse arrays and keys to disk in the aforementioned SequenceFile format.

\subsection*{Mahout K-means clustering}
To cluster the samples from the sequence files, we use Mahout 0.9 with Hadoop 2.5.0. 
The clustering steps are the same for each dataset, however, due to the increased feature size we designate more memory to the
YARN containers. For the first step, we specify \texttt{k},
the number of clusters. The function \texttt{RandomSeedGenerator.buildRandom} creates a new sequence file containing \texttt{k}
random centroids from the samples sequence file. The sequence file
of samples and the sequence file of centroids serve as the input to \texttt{KMeansDriver.run}, a static Mahout API function that launches
a k-means clustering job on the Hadoop service.

\subsection*{R K-means clustering}
We use `readVcf' from the R library, `VariantAnnotation', to iteratively read in the VCF file. For each iteration, we convert the data to a matrix and
convert the values for each variant to a double, as mentioned previously. We then transpose the matrix, so individuals are represented
by rows, and append the matrix to that from the previous iteration. Once the matrix contains all of the variants from the VCF file, we cluster the data
using the `kmeans' algorithm from the `stats' library.

@@

\subsection*{Performance}
We collected the job time from the 'Elapsed' time as reported by the Hadoop JobHistory server. This is measured from the job's start time to finish time.
Because k-means clustering performs multiple iterations, and therefore multiple MapReduce jobs, the time we reported is the sum of the time each iteration consumed.
For the R job, where we could not easily separate the pre-processing and clustering components, we printed the time between
stages using the function, \texttt{proc.time()}.

@@

\begin{table}[!t]
\processtable{Datasets used in this study.\label{Tab:01}}
{\begin{tabular}{lrrrrr}\toprule
& individuals & variants & size (GB) & population\\
& & & &groups& \\\midrule
        \NinteenPhaseone & 1,092 & 816,115 & 7.13  & 4\\
        \OnePhaseone & 1,092 & 3,007,196 & 26.27  & 4\\
        \SevenPhaseone & 1,092 & 18,984,880 & 165.85 & 4\\
	\OnePhasethree\ & 2,535 & 6,468,094 & 131.17 & 5\\
	\ThreePhasethree\ & 2,535 & 19,381,970 & 393.06 & 5\\
	\FullPhasethree\ & 2,535 & 79,000,000 & 1602 & 5\\\botrule
\end{tabular}}{Datasets used in this study; based on portions of the Phase 1 and Phase 3 data from the 1000 genomes project.
Size is the approximate space required for storage of a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
}
\end{table}

\begin{table}[!t]
\processtable{Adjusted Rand index (accuracy).\label{Tab:01}}
{\begin{tabular}{lcc}\toprule
Tool, Dataset & accuracy \\\midrule
        R, \NinteenPhaseone & 0.614  \\ 
        Mahout, \NinteenPhaseone & 0.730\\
        Mahout, \OnePhaseone & 0.891\\
        Mahout, \SevenPhaseone & 0.816 \\
        Mahout, \OnePhasethree & 0.877  \\\botrule
\end{tabular}}{The adjusted Rand index for each dataset. The 'truth' index is the hypothetical cluster-set where there is a cluster for each super-population, with individuals belonging to the same super-population grouped together.}
\end{table}

\begin{table*}[!t]
\processtable{Resources allocated and time taken.\label{Tab:01}}
{\begin{tabular}{lccccc}\toprule
&&\multicolumn{2}{ c }{Memory (GB)}&\multicolumn{2}{ c }{Time} \\
& - & map & reduce  & pre-processing & clustering\\\midrule
         R, \NinteenPhaseone & - & 1.5 & 2 & 55m 41s & 7s\\
        Mahout,\NinteenPhaseone & - & 1.5 & 2 & 2m 59s & 9m 12s\\
        Mahout,\OnePhaseone & - & 1.5 & 2 & 7m 24s & 23m 9s\\
        Mahout, \SevenPhaseone & - & 1.5 & 2 & 47m 43s & - \\ 
        Mahout, \SevenPhaseone & - & 4 & 4 & - & 10h 43m 50s \\ 
        Mahout, \OnePhasethree & - & 1.5 & 2 & 27m 48s & 89m 42s \\
        Mahout, \ThreePhasethree & - & 1.5 & 2 & 82m 14s & - \\
        Mahout, \ThreePhasethree & - & 4 & 4 & - & (in progress) \\
        Mahout, \FullPhasethree & - & 1.5 & 2 & 336m 8s & - \\\botrule
\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.
Where clustering required bigger containers than pre-processing, an additional line is included in the table for clustering, following pre-processing.}
\end{table*}
